|

### 2023-08-15 chapter_4/src/prr/dcr_color_and_non-political_behaviors_prr.qmd
more drafting of front-half of chapter 4 prr

The previous chapters demonstrated that the colors red and blue conveyed information about the partisanship of politicians and about potential discussion partners. The previous chapters also demonstrated that these effects on attitudes also influenced behaviors such as a willingness to vote for the candidate as well as willingness to have a conversation with another about politics. The previous chapter also demonstrated two other pieces to the story: (1) that the group-based information conveyed the colors red and blue are dependent on the context, and (2) that the snap-judgments people form based on this simple visual information can shape attitudes and behaviors even when provided with more clear and explicit information about the partisan affiliation of another person.

This chapter builds on these findings to examine whether these associations between the colors red and blue with the partisanship of another can convey information about groups of people and can guide attitudes and resulting behaviors towards interactions with these groups of people. Specifically, the chapter builds upon a growing set of work suggesting that the politicultural differences between Democrats and Republicans are shaping seemingly non-political attitudes and behaviors such as one's choice in which neighborhood to live in. I argue in this chapter that the prevalence of red and blue cars in the driveways of houses in a neighborhood shift people's perceptions of the partisan composition of the neighborhood and as a result will make people less willing to move to that neighborhood when they perceive they'd be in the political minority of that neighborhood.

@@

While political polarization certainly reflects a growing gap and consolidation of two camps of policy preferences and attitudes [@dellaposta_pluralistic_2020;@luders_attitude_2023], it manifests as something more consequential for our every-day experiences. Political polarization has come to reflect social groups that distinguish "us" versus "them" with increasingly fewer and fewer non-political social groups that are not wrapped up in our partisan group [@mason_uncivil_2018]. That is, we are not only holding fewer and fewer cross-cutting attitudes that moderate our political viewpoints [@luders_attitude_2023], but we also have fewer and fewer cross-cutting social groups that encourage interaction and experiences with perspectives of out-partisans [@mason_uncivil_2018]. The consequence of such phenomena is that we are becoming more and more willing to have strong negative reactions to those of the other political party and strong positive reactions to members of our political party [@iyengar_affect_2012;@iyengar_fear_2015].

The consequences of this affective polarization encourages discomfort being around those who identify with the other political party and finding comfort around those that share the same partisanship. In fact, some evidence suggests that it is even more sinister than comfort but rather predicts lower empathy for outpartisans [@allamong_screw_2021], but also a willingness to describe outpartisans as subhuman [@martherus_party_2021]. To communicate the magnitude of this dislike for out-partisans, dehumanizing language is often used by those that hold deeply racially resentful views [@utych_how_2018].

Evidence of this discomfort with out-partisans is piling up. For example, @carlson_what_2022 demonstrates that individuals perceive that they may incur social costs by engaging in a conversation with those they disagree with and expressing that disagreement. Other work corroborate this finding by suggesting that on social media individuals will be less likely to share the posts of political viewpoints they agree with if they perceive that they are in the political minority of those that would see that sharing of the post [@van_der_does_strategic_2022]. As a result, individuals have fewer outpartisans in their group of folks they regularly discuss politics with [@butters_polarized_2022].

One interesting finding of @butters_polarized_2022 is that one cause of the increasing homogeneity of people's discussion partners' partisanship is that individuals often live in counties  that are politically-homogenous. This finding reflects a growing area of research examining the geographical sorting occurring as a result of polarization among the mass public.

>>>>

The previous chapters demonstrated that the colors red and blue conveyed information about the partisanship of politicians and about potential discussion partners. The previous chapters also demonstrated that these effects on attitudes also influenced behaviors such as a willingness to vote for the candidate as well as willingness to have a conversation with another about politics. The previous chapter also demonstrated two other pieces to the story: (1) that the group-based information conveyed by the colors red and blue are dependent on the context, and (2) that the snap-judgments people form based on this simple visual information can shape attitudes and behaviors even when provided with more clear and explicit information about the partisan affiliation of another person.

This chapter builds on these findings to examine whether these associations between the colors red and blue with the partisanship of another can convey information about groups of people and can guide attitudes and resulting behaviors towards interactions with these groups of people. Specifically, the chapter builds upon a growing set of work suggesting that the politicultural differences between Democrats and Republicans are shaping seemingly non-political attitudes and behaviors such as one's choice in which neighborhood to live in. I argue in this chapter that the prevalence of red and blue cars in the driveways of houses in a neighborhood shift people's perceptions of the partisan composition of the neighborhood, and as a result, will make people less willing to move to that neighborhood when they perceive they'd be in the political minority of that neighborhood.

@@

While political polarization certainly reflects a growing gap and consolidation of two camps of policy preferences and attitudes [@dellaposta_pluralistic_2020;@luders_attitude_2023], it manifests as something more consequential for our every-day experiences. Political polarization has come to reflect social groups that distinguish "us" versus "them" with increasingly fewer and fewer non-political social groups that are not wrapped up in our partisan group [@mason_uncivil_2018]. That is, we are not only holding fewer and fewer cross-cutting attitudes that moderate our political viewpoints [@luders_attitude_2023], but we also have fewer and fewer cross-cutting social groups that encourage interaction and experiences with perspectives of out-partisans [@mason_uncivil_2018]. The consequence of such phenomena is that we are becoming increasingly willing to harbor strong negative feelings towards those of the other political party and strong positive affectations to members of our political party [@iyengar_affect_2012;@iyengar_fear_2015].

The consequences of this affective polarization encourages discomfort being around those who identify with the other political party and finding comfort around those that share the same partisanship. In fact, some evidence suggests that it is even more sinister than comfort but rather predicts lower empathy for outpartisans [@allamong_screw_2021], as well as an increased willingness to describe outpartisans as subhuman [@martherus_party_2021]. Racists commonly use dehumanizing language to describe racial minorities [@utych_how_2018].

Evidence of this discomfort and its manifestations with out-partisans is piling up. For example, @carlson_what_2022 demonstrates that individuals perceive that they may incur social costs with friends and neighbors by engaging in a conversation with those they disagree with and expressing that disagreement, and avoid such situations as a result. Other work corroborate this finding by suggesting that individuals will be less likely to share the social media posts of political viewpoints they agree with if they perceive that they are in the political minority of those that would see that sharing of the post [@van_der_does_strategic_2022]. As a result of this tendency to speak out as a political minority in front of a group of people, individuals have fewer outpartisans in their group of folks they regularly discuss politics with [@butters_polarized_2022].

Building on this set of evidence, a growing area of the literature suggests that we are able to make inferences about the partisanship about other individuals, but we are able to do this for groups of people through the increasing politicultural distinctions between Republicans and Democrats. @hetherington_prius_2018 demonstrates that Republicans and Democrats are not just different in their policy preferences but rather that they live in different places and have different lifestyles. Others have replicated and built upon this work by suggesting that while there are many different types of lifestyles that Republicans and Democrats can take on, there is increasingly clear distinction between Republicans and Democrat archetypes [@rogers_politicultural_2022]. These politicultural distinctions are clear enough that we train a computer to make accurate guesses (upwards of 80%) of the partisanship by using only images of the cars in a particular neighborhood.

Though not inherently or necessarily political, our communities often act in political ways. For example, some evidence suggests despite variation between counties' partisan composition and lockdowns discouraging people from leaving their homes during the COVID-19 pandemic, those living in Republican counties spent more time away from home than those living in Democratic counties[@roberts_polarized_2021]. Further, we see a place-based identity for rural individuals that predict partisanship [@cramer_politics_2016; see also @lyons_youre_2021] and characteristics like expression of anti-intellectual attitudes [@lunz_trujillo_rural_2022]. While place-based identities could be separate or informative of partisan identities, some literature argues that partisanship may actually inform which community we choose to make ours.

The consequences of the fewer cross-cutting social groups and increasingly homogenous cultural preferences of Republicans and Democrats manifest as dramatic differences in preferences about the communities that we are part of. @gimpel_seeking_2015 demonstrates that describing a neighborhood with explicit and implicit information (e.g., racial composition) about the partisan composition influences whether one expresses a desire or not to move and live there. Further work corroborates this finding by providing evidence from a number of studies which demonstrate that not only do people associate particular characteristics of a neighborhood such as the prevalence of churches with Republicans and hybrid cars with Democrats, but also that these features attract copartisans and repulse outpartisans [@motyl_how_2020].

Some work suggests, however, that these expressed preferences to live near copartisans do not carryover to migration patterns [see @iyengar_affect_2012]. In fact, there is some evidence suggesting that controlling for everything else but a political sorting, we would have less political homogeneity than we had in 2008 [@martin_does_2020]. Further, evidence by @martin_does_2020 suggests that people switch their party registration after moving as opposed to moving to be more aligned with their party registration.

The goal of the following sections is to discuss one potential mechanism that may help to tease out why there is weak observational empirical evidence for a dearth of theoretical and empirical evidence
